Optimizing Abstract Abstract Machines

### 2012-11-14 paper/paper.tex
Overhaul of AAM review.

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval   &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\maddr}\\
&&|& \kar{\mexp,\menv,\maddr}\\
&&|& \kif{\mexp,\mexp,\menv,\maddr}
\end{array}
\]
\caption{Abstract machine components}
\end{figure}

Figure~\ref{fig:aam} defines the semantics of ISWIM as a machine
model.  Evaluation is defined as the set of states reachable by the
reflexive, transitive closure of the machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply''
form~\cite{dvanhorn:Danvy:DSc}.  It can be
systematically derived from a definitional interpreter through a
continuation-passing style transformation and defunctionalization, or
from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

Compared with the standard machine semantics, this definition is
different in the following ways:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values,
\item continuations are heap-allocated,
\item there are ``contour values'' (written $\mcntr$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\mathit{push}$, $\mathit{bind}$, $\interpdelta$ and $\sqcup$.
\end{itemize}

@@

\paragraph{Concrete interpretation} can be characterized by setting the implicit
parameters of the relation given in Figure~\ref{fig:aam} as follows:
\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\mathit{bind}(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
\mbox{ where }\maddr \notin\msto
\end{align*}
The resulting relation is non-deterministic in its choice of
addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming, this relation defines
a deterministic machine.  (The relation is really a function.)

\paragraph{Abstract interpretation} can be characterized by setting the implicit
parameters just as above, but dropping the $\maddr \not\in \msto$
condition.  This family of interpreters is also non-deterministic in
choices of addresses, but it is free to choose addresses that are
already in use.  Consequently, the machines may be non-deterministic
when multiple values reside in a store location.

It is important to recognize from this definition that \emph{any}
allocation strategy is an abstract
interpretation~\cite{dvanhorn:Might2009Posteriori}.  In particular,
concrete interpretation is a kind of abstract interpretation.  So is
an interpretation that allocates a single cell into which all bindings
and continuations are stored.  On the one hand is an abstract
interpretation that is non-computable and gives only the ground truth
of a programs behavior; on the other is an abstract interpretation
that is easy to compute but gives little information.  Useful program
analyses lay somewhere in between and can be characterized by their
choice of address representation and allocation strategy.

%% We now have a framework for describing program analysis for the ISWIM
%% family of languages, whereby approximation of both control and
%% environment structure is regulated by the heap and allocation
%% policies.

Uniform \(k\)-CFA is one such analysis.

\paragraph{Uniform \(k\)-CFA} can be characterized by the following allocation
strategy:

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}
$\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto \sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup \msto'(\maddr)$.
For the duration of this paper, since our example abstract value space
has no interesting ordering, $\sqcup$ can be read as $\cup$.
\paragraph{Primitives}

\begin{align*}
\mnum+1 &\in \interpdelta(\saddone,\mnum) &
\mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
\strue &\in \interpdelta(\szerohuh,\szero) &
\sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
\end{align*}

\begin{align*}
\sNum &\in \hat\interpdelta(\saddone,\mnum) &
\sNum &\in \hat\interpdelta(\ssubone,\mnum)\\
\strue &\in \hat\interpdelta(\szerohuh,\sNum) &
\sfalse &\in \hat\interpdelta(\szerohuh,\sNum)\\
\strue &\in \hat\interpdelta(\szerohuh,\szero) &
\sfalse &\in \hat\interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
\end{align*}

%% \begin{figure}
%% \begin{alltt}
%%   ;; State \(\to\) \(\Set(\)State\()\)
%%   (define (step \(\mstate\))
%%     (match \(\mstate\)
%%       [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
%%        (match \(\mexp\)
%%          [(var\(\superscript\mlab\) \(\mvar\))
%%           (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
%%          [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
%%          [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
%%          [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
%%          [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
%%       [(co \(\msto\) \(\mkont\) \(\mval\))
%%        (match \(\mkont\)
%%          ['mt (set (ans \(\msto\) \(\mval\)))]
%%          [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
%%          [(fn\(\superscript\mlab\) \(\mval'\))
%%           (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
%%             (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
%%          [(fi\(\superscript\mlab\) c a \(\rho\))
%%           (for/set ((k (get-cont \(\msto\) l)))
%%             (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
%%       [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
%%        (match fun
%%          [(clos l \(\mvar\) \(\mexp\) \(\rho\))
%%           (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
%%           (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
%%          [(op \(\mop\))
%%           (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
%%                      (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
%%          [_ (set)]))]))
%% \end{alltt}
%% \caption{Implementation of machine transition relation.}
%% \end{figure}

>>>>

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval   &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\maddr}\\
&&|& \kar{\mexp,\menv,\maddr}\\
&&|& \kif{\mexp,\mexp,\menv,\maddr}\\
\text{Environments} &\menv&\in& \mathit{Var} \rightharpoonup \mathit{Addr}\\
\text{Stores} &\msto&\in& \mathit{Addr} \rightharpoonup \mathcal{P}(\mathit{Val})
\end{array}
\]
\caption{Abstract machine components}
\end{figure}

The semantics are defined in terms of a machine model.  The machine
compoents are defined in figure~\ref{fig:domains} and
figure~\ref{fig:aam} defines the transition relation.  The evaluation
of a program is defined as the set of states reachable by the
reflexive, transitive closure of the machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It
can be systematically derived from a definitional interpreter through
a continuation-passing style transformation and defunctionalization,
or from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values, not
  single values,
\item continuations are heap-allocated, not stack-allocated,
\item there are ``contour values'' (written $\mcntr$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\mathit{push}$, $\mathit{bind}$, and $\interpdelta$.
\end{itemize}

@@

\paragraph{Concrete interpretation} To characterize concrete interpretration, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\mathit{bind}_\mlab^\mcntr(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
\mbox{ where }\maddr \notin\msto
\end{align*}
The resulting relation is non-deterministic in its choice of
addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming, this relation defines
a deterministic machine.  (The relation is really a function.)

The interpretation of primitive operations is defined by setting
$\interpdelta$ as follows:
\begin{align*}
\mnum+1 &\in \interpdelta(\saddone,\mnum) &
\mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
\strue &\in \interpdelta(\szerohuh,\szero) &
\sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
\end{align*}

\paragraph{Abstract interpretation} To characterize abstract interpretation, set the implicit
parameters just as above, but drop the $\maddr \not\in \msto$
condition.  This family of interpreters is also non-deterministic in
choices of addresses, but it is free to choose addresses that are
already in use.  Consequently, the machines may be non-deterministic
when multiple values reside in a store location.

It is important to recognize from this definition that \emph{any}
allocation strategy is an abstract
interpretation~\cite{dvanhorn:Might2009Posteriori}.  In particular,
concrete interpretation is a kind of abstract interpretation.  So is
an interpretation that allocates a single cell into which all bindings
and continuations are stored.  On the one hand is an abstract
interpretation that is non-computable and gives only the ground truth
of a programs behavior; on the other is an abstract interpretation
that is easy to compute but gives little information.  Useful program
analyses lay somewhere in between and can be characterized by their
choice of address representation and allocation strategy.  Uniform
\(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97} is one such analysis.

\paragraph{Uniform \(k\)-CFA} To characterize uniform \(k\)-CFA, set the allocation
strategy as follows, for a fixed constant \(k\):

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}
where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto \sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup \msto'(\maddr)$.

All that remains is the interpretation of primitives.  For abstract
intepretation, we set $\interpdelta$ to the function that returns
$\sNum$ on all inputs---a symbolic value we interpret as denote the
set of all integers.

At this point, we have abstracted the original machine to one which
has a finite state space for any given program, and thus forms the
basis of a sound, computable program analyzer for ISWIM.
